---
title: "Data Manipulation Tutorial - Datafest 2019"
author: "Vlad Achimescu"
date: "May 3, 2019"
output: html_document
---

## LOAD PACKAGES

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages(c("readr","magrittr","dplyr","tidyr"))
### load packages
library(readr)
library(magrittr)
library(dplyr)
library(tidyr)
### make sure filter and select are assigned to dplly
filter = dplyr::filter
select = dplyr::select
```


## IMPORT DATA

Read data as a tibble, using read_csv or read_tsv.

Read employment dataset
```{r cars}
EMP = read_tsv("empl_rates_reg.tsv")
EMP
```

Read education&training dataset
```{r}
TRAIN = read_csv("training_participation.csv")
TRAIN
```


## CHAINING FUNCTIONS - PIPE OPERATOR

Pipe operator simplifies reading and writing chained functions.
Shortcut for pipe operator "%>%" is <Ctrl+Shift+M>.
If previous value is not used as a first argument, but later, use "." to specify the position

Example: summarise unemployment rate in 2018

```{r}
# Extract unemployment rates
unempl2018  = 100-EMP$`2018`
summary(unempl2018)

# Base R - chaining
paste("Result:", round(quantile(log(unempl2018,2),.25, na.rm = TRUE), 1))
      
# Base R - separate operations
log_unempl   = log(unempl2018,2)
quant_unempl = quantile(log_unempl, .25, na.rm = TRUE)
round_quant = round(quant_unempl, 1)
paste("Result:", round_quant)

# Magrittr - chaining using pipe
unempl2018 %>% 
  log(2) %>% 
  quantile(.25, na.rm = TRUE) %>% 
  round(1) %>% 
  paste("Result:",.)

# Magrittr - one line
unempl2018 %>% log(2) %>% quantile(.25, na.rm = TRUE) %>% round(1) %>% paste("Result:",.)
```

Create a table of relative frequencies, sorted by descending:
```{r}
table(TRAIN$SEX)
#Base R:

cbind(round(rev(sort(prop.table(table(TRAIN$SEX)))),2))
```

```{r}
### additional functions (aliases) useful for piping
?magrittr::extract
```


## IMPORT DATA

Read data as a tibble, using read_csv or read_tsv.

```{r cars}
EMP = read_tsv("empl_rates_reg.tsv")
EMP
```

```{r}
TRAIN = read_csv("training_participation.csv")
TRAIN
```

## Slice and filter

Slice

```{r pressure, echo=FALSE}
# Base R
TRAIN[1:5,]

# dplyr
slice(TRAIN, 1:5)

# With piping
TRAIN %>% slice(1:5)
```

Subset rows / filter(logical_condition)

```{r}
### Base R - select cases from 2018 with less than 2% training
TRAIN[TRAIN$Value>30 & TRAIN$TIME == 2018,]

### dplyr
TRAIN %>% filter(Value>30, TIME == 2018)
```

Filter by strings - regular expressions / sets

```{r}
### Select all countries starting with the letter "C"
TRAIN %>% filter(grepl("^C",GEO))

### Select Slovakia and Slovenia
TRAIN %>% filter(GEO %in% c("Slovakia","Slovenia"))
```

EXERCISE: select
```{r}
TRAIN %>% filter()
```

## Subset variables/columns - select

Select function makes code similar to Stata

```{r}
### Base R - select first three variables
EMP[,1:3]
EMP[,c("unit", "sex", "age" )]

### dplyr
EMP %>% select(1:3)
EMP %>% select("unit", "sex", "age")
EMP %>% select(unit, sex, age)
EMP %>% select(unit:age)
```

```{r}
### select by skipping
EMP %>% select(geo, age, sex, Y2016:Y2018)

### unselect
EMP %>% select(-unit)

### select by variable name
EMP %>% select(starts_with("Y"))
```

EXERCISE: select 

```{r}
TRAIN %>% select()
```


## Create new variables / mutate, transmute

```{r}
### Base R
TRAIN$low = TRAIN$Value<10

### dplyr
TRAIN %>% mutate(low = Value<10)
```

Mutate keeps all variables, transmute only keeps new variables

```{r}
### mutate
TRAIN %>% mutate(low = Value<10,
                 cumval = cumsum(Value),
                 period = ifelse(TIME<2010, "before 2010","2010 or after"))

### transmute
TRAIN %>% transmute(GEO,
                    low = Value<10,
                    cumval = cumsum(Value),
                    period = ifelse(TIME<2010, "before 2010","2010 or after"))

### can use window functions in : lag, cumsum, min_rank, etc. (see cheat sheet)
```


Mutate specific variables with mutate_if and mutate_at

```{r}
### Change variables that follow a certain condition
## mutate_if(condition, function)
# e.g.make all strings into factors
TRAIN %>% mutate_if(is.character, as.factor)

### Change specific variables
## mutate_at(vars(), funs())
# e.g. round all Y variables
EMP %>% mutate_at(vars(Y2007:Y2018), funs(as.numeric)) %>%
        mutate_at(vars(Y2007:Y2018), funs(round))
```


Mutate_all replaces sapply or tapply

```{r}
### apply same functions to all variables

## apply predefined function
TRAIN %>% mutate_all(nchar)
## apply custom function
TRAIN %>% mutate_all(function(x) paste0(x,"_suffix"))
```


## Aggregate data

```{r}

```

## Reshape datasets

From wide to long format

```{r}

```

From wide to long format

Separate 

```{r}

```

## Combine/merge datasets

Similar to SQL functions

```{r}

```

## All-in-one pipe

```{r}

```
